### 2017-05-10 01-intro.tex
Initial merge of intro, conclusions

In this paper, we focus on the SG-MCMC method applied to the a-MMSB model. Recent work\cite{LiAW15} showed that this methodology is faster and more accurate than the SVB method. \cite{LiAW15} further proposes a heuristic to scale up the number of communities at the cost of less precision; the work in this paper scales the algorithm without that heuristic with a custom high-performance implementation. To this end, we propose a design of a parallel and distributed system specifically tailored to solve the a-MMSB problem. In particular, we use a mixture of OpenMP, MPI and RDMA in order to efficiently scale and accelerate the algorithm's computation.

Achieving this goal necessitated overcoming several challenges. First, the algorithm's state grows rapidly with larger graphs and number of latent communities. Since the full state information is too large to fit in a single machine's memory, it is partitioned and distributed across a cluster of machines. Second, to access the full state, each cluster node must read remote memory hosted by its peers. We leveraged RDMA to limit the high latency of such operations and increase the communication bandwidth. To reduce the latency further, we pipelined the algorithm's computations such that data can be fetched in advance over the network. Finally, the algorithm's computation is effectively distributed across the cluster nodes and parallelized further within each node by exploiting their multi-core CPUs.

The remainder of this paper is organized as follows. Section~2 provides an overview of the algorithm and its theoretical foundation. The design and implementation of the parallel and distributed solution is presented in Section~3. Section~4 evaluates the efficacy of the system and analyzes its performance. Finally, Section~5 provides concluding remarks.

>>>>

In this paper, we focus on the SG-MCMC method applied to the a-MMSB model. Recent
work~\cite{LiAW15} showed that this methodology is faster and more accurate than the
SVB method. \cite{LiAW15} further proposes a heuristic to scale up the number of
communities at the cost of less precision; the work in this paper considers the
algorithm without that heuristic.

From a computational point of view, our SG-MCMC a-MMSB algorithm differs
from widespread machine learning algorithms, like deep learning, in several
ways. First, it is highly data-intensive which makes parallel acceleration
particularly challenging. Second, owing to the algorithm's stochastic
nature, the majority of its memory access patterns and data dependencies
are non-deterministic. As a result, straightforward optimization attempts
of the memory access patterns either fail or lead to non-intuitive results. Third,
the size of its intermediate data structures makes it difficult to scale to very large
communities.

Starting from an existing implementation in Python/Numpy, we will present
how this algorithm can be implemented much more efficiently in C++. Then,
we will describe our two key contributions: first, how to scale up our
implementation to run on GPUs and multicore CPUs, second, to scale out
our implementation to overcome the memory bounds posed by GPUs or general
single-machine systems.
% To this end, we propose a design of a parallel and distributed system specifically tailored to solve the a-MMSB problem. In particular, we use a mixture of OpenMP, MPI and RDMA in order to efficiently scale and accelerate the algorithm's computation.

The focus of the first part of our work is on the computational efficiency and parallel
performance of the SG-MCMC algorithm by
developing aggressive optimizations targeting
multi-core CPUs and GPUs. The GPU implementation achieves speedup factors up to~86,
compared to the
well-tuned sequential C++ program which itself is a factor~1000-1500 faster than
the original Python/Numpy program.
%
Through careful analysis of the computation and data structures we show that
the algorithm's full state can be reduced by roughly 75\%. Compressing the
state significantly reduces the data intensity and allows for tackling larger
problems. % while maintaining all state in memory.

\begin{comment}
Further, by cataloguing and accounting for the various load and store
operations, we identified the highest priority locations of data reuse. In
order to circumvent the unclear optimization landscape, we developed an
effective kernel code generation mechanism that explores the benefits of
exploiting all permutations of the available optimization opportunities. These
optimizations include caching in shared memory, caching in the register file,
loop unrolling and explicit vectorization.
\end{comment}

The second part of our work, scaling out to the largest publicly available
community graphs, also necessitated overcoming several challenges. First,
the algorithm's state grows rapidly with larger graphs and number of latent
communities.  Since the full state for the larger graphs does not fit in a
single machine's memory, it is partitioned and distributed across a cluster
of machines. Second, to access the full state, each cluster node must read
remote memory hosted by its peers. We leveraged RDMA to limit the high latency
of such operations and increase the communication bandwidth. To reduce the
latency further, we pipelined the algorithm's computations such that data can
be fetched in advance over the network. Finally, the algorithm's computation
is effectively distributed across the cluster nodes and parallelized further
within each node by exploiting their multi-core CPUs.

The remainder of this paper is organized as follows.
Section~\ref{sec-background} provides an overview of the
algorithm and its theoretical foundation, with related work in
Section~\ref{sec-related}. Section~\ref{sec-gpu} delves
into the parallelization optimizations for GPUs and multicore CPUs.
Section~\ref{sec-distr} describes the design, implementation and evaluation of
the scalable distributed solution. Finally, Section~\ref{sec-conclusion}
provides concluding remarks.

### 2017-05-10 06-conclusion.tex
Initial merge of intro, conclusions

\section{Conclusion}
Modern machine learning algorithms are empowering their
users to solve or approximate solutions to previously intractable problems.
However, these advancements come at the cost of significant computational
complexity and long running learning processes which hinder our ability to reap
such algorithms' benefits when applied to large problems. There is a clear need
to assess such algorithms and identify optimization opportunities in order to
scale their performance.

@@

We showed that the algorithm's state can be reduced by 75\% after a thorough
analysis of the computational patterns and data structures which significantly
reduces its data intensity. Further, we navigated the complex optimization
landscape by dynamically generating kernel code and testing different
combinations of optimizations. The outcome of these efforts culminated in
significant speedup factors of 21 and 86 using a multi-core CPU and a GPU
respectively. These speedup numbers where achieved in comparison to an already
optimized sequential implementation. Finally, we evaluated the performance of
the parallel algorithm across several GPUs, highlighting the difference between
their optimal configurations.

The outcome of this study reinforces the significance of avoiding premature
optimization as it can lead to unexpected results. In particular, the success
of common GPU optimizations depends on the particular device in use and the
problem it is applied to.

>>>>

\section{Conclusions}
\label{sec-conclusion}

Modern machine learning algorithms are empowering their
users to solve or approximate solutions to previously intractable problems.
However, these advancements come at the cost of significant computational
complexity and long running learning processes which hinder our ability to reap
such algorithms' benefits when applied to large problems. There is a clear need
to assess such algorithms and identify optimization opportunities in order to
scale their performance.

@@

We showed that the algorithm's state can be reduced by 75\% after a thorough
analysis of the computational patterns and data structures which significantly
reduces its data intensity.

In our optimization for manycore processors, we navigated the complex optimization
landscape by dynamically generating kernel code and testing different
combinations of optimizations. The outcome of these efforts culminated in
significant speedup factors of 21 and 86 using a multi-core CPU and a GPU
respectively. These speedup numbers were achieved in comparison to an already
optimized sequential implementation. Finally, we evaluated the performance of
the parallel algorithm across several GPUs, highlighting the difference between
their optimal configurations.
%
The outcome of this study reinforces the significance of avoiding premature
optimization as it can lead to unexpected results. In particular, the success
of common GPU optimizations depends on the particular device in use and the
problem it is applied to.

We further created a highly scalable implementation to handle the largest existing
community graphs.
This design faced
several obstacles in order to achieve high performance.
The algorithm was restructured to facilitate its
parallelization. We overlapped computation
with communication to hide latency.  We use a
mixture of MPI and RDMA primitives to speedup the communication between cluster
nodes.
%
Further, we conducted a thorough empirical evaluation of the system to study its strong
and weak scalability on 65 cluster nodes using large data sets.
% Additionally, we assessed the efficiency of the algorithm's resource utilization.
Finally, a
demonstration of the implementation's utility was provided by processing 6
different organic data sets.
To the best of
our knowledge, this is the first time that the problem of deducing overlapping
communities has been learned for problems of such a large scale.
